---
title: "Data Scientist A - Task 1: Merge Example Data"
author: "Mikhaela Lewis"
date: "2025-07-05"
output: 
  html_document:
    theme: darkly
    toc: true
    toc_depth: 3
---
# Task 1: Merge Example Data


## Setup


### Setup Knitr

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Attach libraries

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
#library(purrr)
```


#### Save filepaths

```{r}
file1 <- '/Users/mikhaelalewis/Documents/ai4purpose-25/dsa/set1_20240101-20240530.xlsx'
file2 <- '/Users/mikhaelalewis/Documents/ai4purpose-25/dsa/set2_20240701-20241231.xlsx'
file3 <- '/Users/mikhaelalewis/Documents/ai4purpose-25/dsa/set3_20250101-20250528.xlsx'
```


#### Function to create a named list of dataframes for each sheet in an Excel file

```{r}
get_sheets <- function(file_name) {
  # Get all sheet names
  sheet_names <- readxl::excel_sheets(file_name)
  
  # Read each sheet into a named list of dataframes
  sheet_list <- lapply(sheet_names, function(one_sheet) as.data.frame(readxl::read_excel(file_name, sheet = one_sheet)))
  names(sheet_list) <- sheet_names
  
  return(sheet_list)
}
```


#### Use functions on the 3 Excel files

```{r}
set1_dfs <- get_sheets(file1)
set2_dfs <- get_sheets(file2)
set3_dfs <- get_sheets(file3)
```


### Check loading

```{r}
# Ensure that each named list of dataframes has the same names
names(set1_dfs)
names(set2_dfs)
names(set2_dfs)
```


### Ensure in tabular data are dataframes, not tibbles

```{r}
class(set1_dfs[["Heart Rate"]]) # df
class(set2_dfs[["Heart Rate"]]) # df
class(set3_dfs[["Heart Rate"]]) # df
```


## Cleaning: Column Names


### Reset Code

```{r}
# For resetting df lists between test runs

set1_dfs <- get_sheets(file1)
set2_dfs <- get_sheets(file2)
set3_dfs <- get_sheets(file3)
```


### Rename columns and convert data types where necessary


#### 1. Heart Rate


##### Create copies

```{r}
hr1 <- set1_dfs[["Heart Rate"]] # Unedited copy of Heart Rate df from Set 1
hr1_test1 <- hr1 # copy for performing 1st test
```


##### Basic Logic

The following basic logic works.

```{r}
# Rename then retype

# Column names
old <- c("Date", "Min", "Max", "Avg", "Median")
new <- c("date", "hr_min", "hr_max", "hr_avg", "hr_med")

hr1_test1 <- hr1_test1 %>% rename_with(~ new, all_of(old)) # Rename columns

# Retyping columns -> all numeric except for date, which is a Date
hr1_test1 <- hr1_test1 %>% 
  mutate(
    date = mdy(date),
    across(new[- 1], as.numeric)
  )

# Check changes
str(hr1_test1)
```

##### Combined Logic

The combined logic works also.

```{r}
hr1_test2 <- hr1 # make copy for 2nd test

# Rename and retype in one step

# Column names
old <- c("Date", "Min", "Max", "Avg", "Median")
new <- c("date", "hr_min", "hr_max", "hr_avg", "hr_med")

# Apply changes 
hr1_test2 <- hr1_test2 %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    across(new[- 1], as.numeric)
  )

# Check changes
str(hr1_test2)
```

##### Function

```{r}
reshape_hr <- function(df) {
  # Column names
  old <- c("Date", "Min", "Max", "Avg", "Median")
  new <- c("date", "hr_min", "hr_max", "hr_avg", "hr_med")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    across(new[- 1], as.numeric)
  )
  
  return(df)
}
```


##### Test on Set 1 Heart Rate data

The function works on Set 1 Heart Rate data.

```{r}
hr1_test3 <- hr1 # make copy for 3rd test
hr1_test3 <- reshape_hr(hr1_test3) # apply changes
str(hr1_test3) # check changes
```

##### Test on Set 2 Heart Rate data

The function works for Set 2 Heart Rate data.

```{r}
hr2 <- set2_dfs[["Heart Rate"]] # Unedited copy of Heart Rate df from Set 2
hr2_test3 <- hr2 # make copy for test
hr2_test3 <- reshape_hr(hr2_test3) # apply changes
str(hr2_test3) # check changes
```

##### Test on Set 3 Heart Rate data

The function works for Set 3 Heart Rate data.

```{r}
hr3 <- set3_dfs[["Heart Rate"]] # Unedited copy of Heart Rate df from Set 2
hr3_test3 <- hr3 # make copy for test
hr3_test3 <- reshape_hr(hr3_test3) # apply changes
str(hr3_test3) # check changes
```

#### Rename and retype columns in Heart Rate data frames in original named lists

```{r}
# Apply chnages
set1_dfs[["Heart Rate"]] <- reshape_hr(set1_dfs[["Heart Rate"]])
set2_dfs[["Heart Rate"]] <- reshape_hr(set2_dfs[["Heart Rate"]])
set3_dfs[["Heart Rate"]] <- reshape_hr(set3_dfs[["Heart Rate"]])

# Check changes
str(set1_dfs[["Heart Rate"]])
str(set2_dfs[["Heart Rate"]])
str(set3_dfs[["Heart Rate"]])
```

#### 2. Breathing Rate


##### Create copies

```{r}
br1 <- set1_dfs[["Breathing Rate"]] # Unedited copy of Heart Rate df from Set 1
br1_test1 <- br1 # copy for performing 1st test
```


##### Function

```{r}
reshape_br <- function(df) {
  # Column names
  old <- c("Date", "Min", "Max", "Avg", "Median")
  new <- c("date", "br_min", "br_max", "br_avg", "br_med")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    across(new[- 1], as.numeric)
  )
  
  return(df)
}
```


##### Rename and retype columns in Breathing Rate data frames in original named lists

```{r}
# Apply changes
set1_dfs[["Breathing Rate"]] <- reshape_br(set1_dfs[["Breathing Rate"]])
set2_dfs[["Breathing Rate"]] <- reshape_br(set2_dfs[["Breathing Rate"]])
set3_dfs[["Breathing Rate"]] <- reshape_br(set3_dfs[["Breathing Rate"]])

# Check changes
str(set1_dfs[["Breathing Rate"]])
str(set2_dfs[["Breathing Rate"]])
str(set3_dfs[["Breathing Rate"]])
```


#### 3. Oxygen Level

##### Function

```{r}
reshape_ox <- function(df) {
  # Column names
  old <- c("Date", "Min", "Max", "Avg", "Median")
  new <- c("date", "ox_min", "ox_max", "ox_avg", "ox_med")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    across(new[- 1], as.numeric)
  )
  
  return(df)
}
```


##### Rename and retype columns in Oxygen Level data frames in original named lists

```{r}
# Apply changes
set1_dfs[["Oxygen Level"]] <- reshape_ox(set1_dfs[["Oxygen Level"]])
set2_dfs[["Oxygen Level"]] <- reshape_ox(set2_dfs[["Oxygen Level"]])

# Check changes
str(set1_dfs[["Oxygen Level"]])
str(set2_dfs[["Oxygen Level"]])
```


#### Sleep


##### Function

```{r}
reshape_sleep <- function(df) {
  # Column names
  old <- c("Date", "Light", "Deep", "Rem", "Awake", "Total")
  new <- c("date", "sleep_light", "sleep_deep", "sleep_rem", "sleep_awake", "sleep_total")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    across(new[- 1], parse_number)
  )
  
  return(df)
}
```


##### Rename and retype columns in Sleep data frames in original named lists

```{r}
# Apply changes
set1_dfs[["Sleep"]] <- reshape_sleep(set1_dfs[["Sleep"]])
set2_dfs[["Sleep"]] <- reshape_sleep(set2_dfs[["Sleep"]])
set3_dfs[["Sleep"]] <- reshape_sleep(set3_dfs[["Sleep"]])

# Check changes
str(set1_dfs[["Sleep"]])
str(set2_dfs[["Sleep"]])
str(set3_dfs[["Sleep"]])
```


#### Sleep Activities

##### Function

```{r}
reshape_sleepact <- function(df) {
  # Column names
  old <- c("Date", "Description", "Is Important", "Time", "Created By")
  new <- c("date", "sleepact_desc", "sleepact_imp", "sleepact_time", "sleepact_cb")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    sleepact_imp = as.factor(sleepact_imp),
    sleepact_time = parse_time(sleepact_time),
    sleepact_cb = as.factor(sleepact_cb)
  )
  
  return(df)
}
```


##### Rename and retype columns in Sleep Activities data frames in original named lists

```{r}
# Apply changes
set1_dfs[["Sleep Activities"]] <- reshape_sleepact(set1_dfs[["Sleep Activities"]])
set2_dfs[["Sleep Activities"]] <- reshape_sleepact(set2_dfs[["Sleep Activities"]])
set3_dfs[["Sleep Activities"]] <- reshape_sleepact(set3_dfs[["Sleep Activities"]])

# Check changes
str(set1_dfs[["Sleep Activities"]])
str(set2_dfs[["Sleep Activities"]])
str(set3_dfs[["Sleep Activities"]])
```


#### Activities

##### Function

```{r}
reshape_act <- function(df) {
  # Column names
  old <- c("Date", "Description", "Is Important", "Time", "Category", "Created By")
  new <- c("date", "act_desc", "act_imp", "act_time", "act_cat", "act_cb")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    act_imp = as.factor(act_imp),
    act_time = parse_time(act_time),
    act_cat = as.factor(act_cat),
    act_cb = as.factor(act_cb)
  )
  
  return(df)
}
```


##### Rename and retype columns in Activities data frames in original named lists

```{r}
# Apply changes
set1_dfs[["Activities"]] <- reshape_act(set1_dfs[["Activities"]])
set2_dfs[["Activities"]] <- reshape_act(set2_dfs[["Activities"]])
set3_dfs[["Activities"]]  <- reshape_act(set3_dfs[["Activities"]])

# Check changes
str(set1_dfs[["Activities"]])
str(set2_dfs[["Activities"]])
str(set3_dfs[["Activities"]])
```


#### Observations

##### Function

```{r}
reshape_obs <- function(df) {
  # Column names
  old <- c("Date", "Description", "Is Important", "Time", "Created By")
  new <- c("date", "obs_desc", "obs_imp", "obs_time", "obs_cb")

  # Rename and retype in one step
  df <- df %>%
  # Rename columns
  rename_with(~ new, all_of(old)) %>%
  # Retype columns
  mutate(
    date = mdy(date),
    obs_imp = as.factor(obs_imp),
    obs_time = parse_time(obs_time),
    obs_cb = as.factor(obs_cb)
  )
  
  return(df)
}
```


##### Rename and retype columns in Observations data frames in original named lists

```{r}
# Apply changes
set1_dfs[["Observations"]] <- reshape_obs(set1_dfs[["Observations"]])
set2_dfs[["Observations"]] <- reshape_obs(set2_dfs[["Observations"]])
set3_dfs[["Observations"]] <- reshape_obs(set3_dfs[["Observations"]])

# Check changes
str(set1_dfs[["Observations"]])
str(set2_dfs[["Observations"]])
str(set3_dfs[["Observations"]])
```


## Merging: Within Sets


### Helper Function: join_dfs()

Takes a named list of data frames and a numeric representing the set number of the list (a value from 1-3) and returns a data frame resulting from nested left joins of the data frames in the given list.

```{r}
join_dfs <- function(df_list, set_num) {
  # Complete nest joins
  all_joined <- df_list[["Heart Rate"]] %>%
    left_join(df_list[["Sleep"]], by="date") %>%
    left_join(df_list[["Sleep Activities"]], by="date") %>%
    left_join(df_list[["Activities"]], by="date") %>%
    left_join(df_list[["Breathing Rate"]], by="date") %>%
    left_join(df_list[["Observations"]], by="date")
  
  # Add "Oxygen Level" for sets 1-2
  if (set_num == 1 | set_num == 2) {
    all_joined <- all_joined %>%
      left_join(df_list[["Oxygen Level"]], by="date")
  }
  
  return(all_joined)
}
```


### Joining Set 1 Data

```{r}
set1_joined <- join_dfs(set1_dfs, 1) # Perform joins
head(set1_joined) # Check first 6 rows of joined table
tail(set1_joined) # Check last 6 rows of joined table
```

The join was successful, but there are multiple rows per date because one date can have multiple activity, observation, and/or sleep activity descriptions. At this point, the joined table can be kept in long format until we begin building the summarizer and report generator. Once that is underway, the data should be widened so that each row contains all of the descriptions for various activities etc. without a need to make multiple rows corresponding to a single date.


### Joining Set 2 Data

```{r}
set2_joined <- join_dfs(set2_dfs, 2) # Perform joins
head(set2_joined) # Check first 6 rows of joined table
tail(set2_joined) # Check last 6 rows of joined table
```


### Joining Set 3 Data

```{r}
set3_joined <- join_dfs(set3_dfs, 3) # Perform joins

# Add blank oxygen level columns 
set3_joined$ox_min <- as.numeric(rep(NA, nrow(set3_joined)))
set3_joined$ox_max <- as.numeric(rep(NA, nrow(set3_joined)))
set3_joined$ox_avg <- as.numeric(rep(NA, nrow(set3_joined)))
set3_joined$ox_med <- as.numeric(rep(NA, nrow(set3_joined)))

head(set3_joined) # Check first 6 rows of joined table
tail(set3_joined) # Check last 6 rows of joined table
```


## Merging: All Sets Together

Since the joined data frames for each set 1-3 all have the same column names and data types, we can combine the data frames together using a row bind instead of joins.

```{r}
# Perform row bind
all_joined <- bind_rows(set1_joined, set2_joined, set3_joined)

head(all_joined) # Check first 6 rows of joined table
tail(all_joined) # Check last 6 rows of joined table

```

## Cleaning: Filling/Flagging Missing Values

Get column names and data types for check
```{r}
str(all_joined)
```

### Filling/Flagging Summary

- The following columns contain data that is key for generating timeline-based summaries. As a result, missing values for these columns will be flagged and not imputed:
  - *_time -> time that an event was logged
  - *_desc -> description of the event
  - *_imp -> whether the logging clinician marked the event as significant
  - *_cb -> anonymized code for the logging clinician
  
- The following columns contain numeric vital data. Missing columns for these columns will be left as-is and filled only for visualization generation. Light impoutation may be applied for plotting.:
  - *_min -> minimum
  - *_max -> maximum
  - *_avg -> average
  - *_med -> median
  - *_total -> total
  
- The following columns contain data on Patient X's sleep stages. Missing values for these columns will be flagged, and may be aggregated for creating summaries in the future:
  - sleep_light -> hours of light sleep
  - sleep_deep -> hours of deep sleep
  - sleep_rem -> hours of REM sleep
  - sleep_total -> total hours of sleep
  
- The following columns contain data on activities during the day/night that Patient X engaged in, or observations that a clinician made of them. Missing values for these columns will be flagged.
  - act_desc -> description of an activity that Patient X engaged in
  - act_cat -> category of an activity
  - sleepact_desc -> description of a sleep-related event that occurred
  - obs_desc -> description of an observation made by a clinician
  
### Missingness Flags

```{r}
#library(dplyr)

# Mutate merge df with missingness flags
all_joined <- all_joined %>%
  mutate(
    act_desc_missing  = is.na(act_desc),
    act_imp_missing   = is.na(act_imp),
    act_time_missing  = is.na(act_time),
    sleep_desc_missing = is.na(sleepact_desc),
    sleep_imp_missing  = is.na(sleepact_imp),
    sleep_time_missing = is.na(sleepact_time),
    obs_desc_missing  = is.na(obs_desc),
    obs_imp_missing   = is.na(obs_imp),
    obs_time_missing  = is.na(obs_time),
    obs_cb_missing    = is.na(obs_cb)
  )

```



```{r}
# Check for missingness flags
str(all_joined) # flags applied successfully
```


### Clean Key Fields (for LLM use)

```{r}
all_joined <- all_joined %>%
  mutate(
    act_imp_clean       = ifelse(is.na(act_imp), "Unknown", as.character(act_imp)),
    sleepact_imp_clean = ifelse(is.na(sleepact_imp), "Unknown", as.character(sleepact_imp)),
    obs_imp_clean       = ifelse(is.na(obs_imp), "Unknown", as.character(obs_imp)),
    
    act_desc_clean      = ifelse(is.na(act_desc), "No activity recorded", act_desc),
    sleepact_desc_clean = ifelse(is.na(sleepact_desc), "No sleep-related activity recorded", sleepact_desc),
    obs_desc_clean      = ifelse(is.na(obs_desc), "No observation was recorded", obs_desc)
  )

```


```{r}
# Check for replaced NA values with appropriate fillings
str(all_joined)
```


## Export Merged/Cleaned Data

```{r}
saveRDS(all_joined, file = "clean_patient_x.rds")
```

